---
title: "Preparing species list"
author: "Fleur Petersen"
date: "2023-11-06"
output: html_document
---

Preparation RadIUS species list
1. Comparison to the Union list dataset (Adriaens & Oldoni, 2023)
2. Updating ANB species selection
3. Adding RIPARIAS alert list species
4. Checking GBIF codes 


# Setup
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r load_libraries}
library(tidyverse)   
library(rgbif)        
library(trias)
```

```{r clear_memory}
rm(list=ls())
```

# 1. Comparison PrIUS list and Union list

## Get PrIUS species list and Union list
```{r prius_list}
prius_list <- read_csv("./prius/data/input/prius_species_list.csv", locale = locale(encoding = "ISO-8859-1"), show_col_types = FALSE)  
```

```{r union_list}
datasetKey <- "79d65658-526c-4c78-9d24-1870d67f8439" #linked to unionlist dataset ("https://www.gbif.org/dataset/79d65658-526c-4c78-9d24-1870d67f8439")
union_list <- name_usage(datasetKey = datasetKey, limit = 10000)$data %>%
  subset(rank != "KINGDOM") #Animalia, Chromista, Plantae wegfilteren
```

## Small adjustments of species names in both datasets to assure comparability
```{r union_list_adjustments}
union_list$scientificName[union_list$scientificName == "Sciurus carolinensis Gmelin,1788"] <- "Sciurus carolinensis Gmelin, 1788"
union_list$scientificName[union_list$scientificName == "Sciurus niger Linnaeus, 1758"] <- "Sciurus niger L., 1758"
union_list$scientificName[union_list$scientificName == "Nasua nasua Linnaeus, 1766"] <- "Nasua nasua L., 1766"
union_list$scientificName[union_list$scientificName == "Procyon lotor Linnaeus, 1758"] <- "Procyon lotor L., 1758"
union_list$scientificName[union_list$scientificName == "Salvinia molesta D.S. Mitch."] <- "Salvinia molesta D.S.Mitch."
union_list$scientificName[union_list$scientificName == "Lysichiton americanus Hultén and St. John"] <- "Lysichiton americanus Hultén & H. St. John"
```

```{r prius_list_adjustments}
prius_list$Species[prius_list$Species == "Triadica sebífera (L.) Small"] <- "Triadica sebifera (L.) Small"
```

## Update GBIF scientific name in both lists
```{r update-scientific-names}
prius_list$prius_scientificName <- 0

for(i in 1:nrow(prius_list)) {
  name_usage(key = prius_list$GBIF_code[i]) -> temp
  temp$data$scientificName -> prius_list$prius_scientificName[i]
  }

union_list$union_scientificName <- 0

for(i in 1:nrow(union_list)) {
  name_usage(key = union_list$nubKey[i]) -> temp
  temp$data$scientificName -> union_list$union_scientificName[i]
  }
rm(i,temp)
```

## Combining both lists for comparison
```{r joining_species_lists}
full_join(prius_list, union_list, by = c("Species" = "scientificName")) %>%
  transmute(Species, Soort, Lijst, prius_gbif_code = GBIF_code, prius_scientificName, union_gbif_code = nubKey, union_scientificName) %>%
  mutate(
    code_check = if_else(prius_gbif_code == union_gbif_code, "TRUE", "FALSE"),
    prius_name_check = if_else(Species == prius_scientificName, "TRUE", "FALSE")) -> list_check
```

## Check taxonomic status and whether any codes were recently deleted from GBIF
```{r check-taxonomic-status-and-deleted}
# Define a function to check the taxonomicStatus and whether the GBIF code was (recently) deleted
check_status_and_deleted <- function(code_column, status_column, deleted_column) {
  result <- lapply(code_column, function(code) {
    if (is.na(code)) {
      status <- "NA"
      deleted <- "NA"
    } else {
      name_usage(key = code) -> temp
      status <- temp$data$taxonomicStatus
      deleted <- if ("deleted" %in% colnames(temp$data)) "TRUE" else "FALSE"
    }
    return(data.frame(Status = status, Deleted = deleted))
  })
  return(do.call(rbind, result))
}

# Check the status and deleted columns for Prius list
prius_result <- check_status_and_deleted(list_check$prius_gbif_code, "prius_status_check", "prius_deleted")

# Check the status and deleted columns for Union list
union_result <- check_status_and_deleted(list_check$union_gbif_code, "union_status_check", "union_deleted")

# Assign the results back to the data frame
list_check$prius_status_check <- prius_result$Status
list_check$union_status_check <- union_result$Status
list_check$prius_deleted <- prius_result$Deleted
list_check$union_deleted <- union_result$Deleted

# Remove unnecessary variables
rm(prius_result, union_result)
```

## Add issue column that separates different issues
```{r add-issue-column}
list_check <- list_check %>%
  mutate(issue = case_when(code_check == "TRUE" & prius_name_check == "TRUE" ~ "ok",
                           code_check == "TRUE" & prius_name_check == "FALSE" ~ "name_issue",   #name_issue: in prius list Species is not equal to gbif scientific name
                           code_check == "FALSE" & prius_name_check == "TRUE" ~ "code_issue",   #code_issue: gbif codes differ between prius and union list
                           code_check == "FALSE" & prius_name_check == "FALSE" ~ "code_name_issue",   #code_name_issue: combination of two previous issues
                           is.na(code_check) & prius_name_check == "FALSE" ~ "name_issue")) %>%
  arrange(issue)
```

```{r export-list-check}
#write.csv2(list_check, "./radius/data/input/prius_species_list_check.csv")
```


## Adjusting PrIUS list based on Union list check
```{r adjusting-hydrocotyle-ranunculoides-GBIF-code}
radius_list <- prius_list %>%
  mutate(GBIF_code = if_else(GBIF_code == 3034613, 7978544, GBIF_code))  %>%  #aanpassing obv check met union list
  select(Groep, Species, Soort, Lijst, GBIF_code) 
```

# 2. Updating ANB species selection

## Weglaten niet-relevante soorten: Aziatische tijgermug, alsemambrosia
```{r removing-non-relevant-species}
radius_list <- radius_list %>%
  subset(!Soort %in% c("Aziatische tijgermug", "alsemambrosia") )
```

## Toevoegen Sikahert (2440954), Chinese moerasslak (9738098), Chinese prachtslang (8889878), wilde rijst (10919373)
```{r expanding-anb-species-selection}
ANB_extra <- data.frame(matrix(ncol=5,nrow=4, dimnames=list(NULL, c("Groep","Species","Soort","Lijst","GBIF_code")))) %>%
  mutate(GBIF_code = c("2440954","9738098", "8889878", "10919373"))

for(i in 1:nrow(ANB_extra)) {
  name_usage(key = ANB_extra$GBIF_code[i]) -> temp
  temp$data$kingdom -> ANB_extra$Groep[i]
  temp$data$scientificName -> ANB_extra$Species[i]
}
rm(i,temp)

ANB_extra <- ANB_extra %>%
  mutate(Groep = ifelse(Groep == 'Plantae',"plant","dier"), 
         Lijst = "ANB", 
         Soort = c("Sikahert", "Chinese moerasslak", "Chinese prachtslang", "Mantsjoerese wilde rijst")) 
```

## ANB uitbreiding toevoegen aan radius lijst
```{r combining-anb-radius}
radius_list_update <- rbind(radius_list, ANB_extra)
```

# 3. Adding RIPARIAS alert list species

```{r adding-riparias-alert-species}
datasetKey <- "fd004d9a-2ea4-4244-bb60-0df508d20a15" #linked to RIPARIAS alert list ("https://www.gbif.org/dataset/fd004d9a-2ea4-4244-bb60-0df508d20a15")
riparias_list <- name_usage(datasetKey = datasetKey)$data %>%
  subset(rank != "KINGDOM")

riparias_species <- riparias_list[!riparias_list$nubKey %in% radius_list_update$GBIF_code,] %>%   #only select species that are not already present in the radius_list_update
  transmute(Groep = ifelse(kingdom == 'Plantae',"plant","dier"), Species = scientificName, Soort = vernacularName, Lijst = "RIPARIAS", GBIF_code = nubKey) %>%
  mutate(Soort = case_when(Soort == "Kentucky river crayfish" ~ "Kentuckyrivierkreeft",
                           Soort == "calico crayfish" ~ "Calicotrivierkreeft", 
                           Soort == "white river crayfish" ~ "Gestreepte Amerikaanse rivierkreeft",
                           Soort == "Australian yabbi" ~ "Gewone yabby",
                           Soort == "Chinese lizard tail" ~ "Moerasanemoon",
                           Soort == "Cape-pondweed" ~ "Kaapse waterlelie", 
                           Soort == "greater pondweed" ~ "Egeria",
                           Soort == "pickerelweed" ~ "Moerashyacint",
                           Soort == "giant butterbur" ~ "Japans hoefblad", 
                           Soort == "lizard's tail" ~ "Leidse plant",
                           Soort == "monkeyflower" ~ "Gele maskerbloem",
                           Soort == "Manchurian wildrice" ~ "Mantsjoerese wilde rijst"
  ))
```

## Combining radius list and RIPARIAS species
```{r combining-radius-riparias}
radius_list_update <- rbind(radius_list_update, riparias_species)
```

# 4. verifying GBIF keys against GBIF Backbone

```{r radius_list_key_check}
radius_key_check <- gbif_verify_keys(radius_list_update, col_keys = "GBIF_code") %>%
  mutate(key = as.character(key)) %>%
  full_join(radius_list_update, by = c("key" = "GBIF_code"))
```

# add GBIF scientific name as a column
```{r add-gbif-scientific-name}
radius_list_update$GBIF_scientific_name <- 0

for(i in 1:nrow(radius_list_update)) {
  name_usage(key = radius_list_update$GBIF_code[i]) -> temp
  temp$data$scientificName -> radius_list_update$GBIF_scientific_name[i]}
rm(i,temp)
```

# Export RadIUS species list

```{r export-radius-list}
write.csv(radius_list_update, "./radius/data/input/radius_species_list.csv", row.names = FALSE)
```

