Version: 2025-08-11

The species list of PrIUS-II is a compilation of

  1. the (previous) PrIUS list
  2. the fourth Union list update (fifth list)
  3. LIFE RIPARIAS
  4. LIFE DUNIAS
  5. Technisch Vademecum
  6. ad hoc selection

```{r prep packages}
library(tidyverse)
library(readr)
library(rgbif)
```

```{r prep memory}
rm(list = ls())
rm(list = setdiff(ls(), c("raw", "clean")))
```

# PrIUS

Largely inspired on the RadIUS workflow (Fleur): we will take the PrIUS list, but let us first check whether it complies with the Union list (pre-update), and whether nothing has changed on GBIF. 

```{r zenodo}
# For reference
# Note : link retrieved by rightclicking on 'Download' button in Zenodo
zenodo <- read.csv(
  "https://zenodo.org/records/7678524/files/Dhondt_etal_2022_PrIUS_table.csv?download=1"
  )
```

```{r prius}
prius_raw <- read_csv("./prius/data/input/prius_species_list.csv",
                       locale = locale(encoding = "ISO-8859-1"),
                       show_col_types = FALSE)

prius <- prius_raw %>%
  filter(Lijst %in% c("1", "2", "3", "4")) %>% 
  rename(EU_legal_name = Species)  # To clarify where this name comes from # IS HET NIET !!
```

```{r union}
# dataset "https://www.gbif.org/dataset/79d65658-526c-4c78-9d24-1870d67f8439"
union_raw <- name_usage(datasetKey = "79d65658-526c-4c78-9d24-1870d67f8439",
                    limit = 10000)$data %>%
  subset(rank != "KINGDOM") # Animalia, Chromista, Plantae wegfilteren

union <- union_raw
```

Small corrections of species names, as based on the Implementing Regulations, in both datasets to assure comparability.

```{r species name corrections}

## mistakes seemingly on behalf of GBIF dataset
## note : issue raised as https://github.com/riparias/union-list/issues/21 - may need follow-up
union$scientificName[union$scientificName == "Sciurus carolinensis Gmelin,1788"] <-
  "Sciurus carolinensis Gmelin, 1788"
union$scientificName[union$scientificName == "Salvinia molesta D.S. Mitch."] <-
  "Salvinia molesta D.S.Mitch."
union$scientificName[union$scientificName == "Triadica sebifera (L.) Small"] <-
  "Triadica sebífera (L.) Small"

## mistakes seemingly on behalf of prius
## ref. : https://eur-lex.europa.eu/legal-content/EN/TXT/HTML/?uri=CELEX:32016R1141&qid=1754922548880
prius$EU_legal_name[prius$EU_legal_name == "Sciurus niger L., 1758"] <-
  "Sciurus niger Linnaeus, 1758"
prius$EU_legal_name[prius$EU_legal_name == "Nasua nasua L., 1766"] <-
  "Nasua nasua Linnaeus, 1766"
prius$EU_legal_name[prius$EU_legal_name == "Procyon lotor L., 1758"] <-
  "Procyon lotor Linnaeus, 1758"
prius$EU_legal_name[prius$EU_legal_name == "Lysichiton americanus Hultén & H. St. John"] <-
   "Lysichiton americanus Hultén and St. John"
```

Update GBIF scientific names in prius. Make sure to be logged in to GBIF!

```{r}
# ditch the previous names
prius <- prius %>% 
  select(-GBIF_scientific_name)

# go get the names
prius$prius_scientificName <- 0

for(i in 1:nrow(prius)) {
  name_usage(key = prius$GBIF_code[i]) -> temp
  temp$data$scientificName -> prius$prius_scientificName[i]
}

rm(i,temp)
```

For optimal comparison, let's do the same in union.

```{r update-scientific-names}
union$union_scientificName <- 0
for(i in 1:nrow(union)) {
  name_usage(key = union$nubKey[i]) -> temp
  temp$data$scientificName -> union$union_scientificName[i]
}
rm(i,temp)
```

Compare the two lists.

```{r join prius-union}
list_compare <- full_join(prius, union, by = c("prius_scientificName" = "union_scientificName")) %>%
  transmute(scientificName = prius_scientificName,
            prius_gbif_code = GBIF_code,
            union_nubKey = nubKey,
            ) %>%
  arrange(scientificName) %>% 
  mutate(code_check = if_else(prius_gbif_code == union_nubKey, "ok", ""))
```

Here, I prefer a manual check.

As of 2025-08-11, 12 lines give a difference.

1 prius lists Cenchrus setaceus (Forssk.) Morrone 5828232
2 union lists Pennisetum setaceum (Forssk.) Chiov. 2706134
CHECK : 2 is synonym of 1 - no fix needed

1 prius lists Heracleum persicum Desf. 3628745
2 union lists Heracleum persicum Desf. ex Fisch., C.A.Mey. & Avé-Lall.	8000520
CHECK : 2 is synonym of 1 - no fix needed

1 prius lists Pontederia crassipes Mart. 2765942
2 union lists Eichhornia crassipes (Mart.) Solms 2765940
CHECK : 2 is synonym of 1 - no fix needed

1 union lists Rana catesbeiana Shaw, 1802 2427092
2 prius doesn't
CHECK : union listed this as a synonym on top of Lithobates c. - no fix needed

1 prius lists Vespa velutina Lepeletier, 1836 1311477
2 union doesn't
CHECK : prius does to make sure to take all V. velutina, although only nigrithorax is listed - no fix needed

1 prius lists Lampropeltis getula (Linnaeus, 1766) 9799308
2 union lists Lampropeltis getula (Blainville, 1835) 5224480
CHECK : 2 seems to be a doubtful taxon - FIX NEEDED

1 Hydrocotyle ranunculoides L.f. 3034613
2 Hydrocotyle ranunculoides L.fil. 7978544
CHECK : 1 has been deleted - FIX NEEDED

##########

Check taxonomic status and whether any codes were recently deleted from GBIF.

```{r make function}
# Define a function to check the taxonomicStatus and whether the GBIF code was (recently) deleted
check_status_and_deleted <- function(code_column, status_column, deleted_column) {
  result <- lapply(code_column, function(code) {
    if (is.na(code)) {
      status <- "NA"
      deleted <- "NA"
    } else {
      name_usage(key = code) -> temp
      status <- temp$data$taxonomicStatus
      deleted <- if ("deleted" %in% colnames(temp$data)) "TRUE" else "FALSE"
    }
    return(data.frame(Status = status, Deleted = deleted))
  })
  return(do.call(rbind, result))
}
```

```{r use function}
# Check the status and deleted columns for Prius list
prius_result <- check_status_and_deleted(list_check$prius_gbif_code,
                                         "prius_status_check", "prius_deleted")

# Check the status and deleted columns for Union list
union_result <- check_status_and_deleted(list_check$union_gbif_code,
                                         "union_status_check", "union_deleted")

# Assign the results back to the data frame
list_check$prius_status_check <- prius_result$Status
list_check$union_status_check <- union_result$Status
list_check$prius_deleted <- prius_result$Deleted
list_check$union_deleted <- union_result$Deleted

# Remove unnecessary variables
rm(prius_result, union_result)
```

Add issue column that separates different issues.

```{r add-issue-column}
list_check <- list_check %>%
  mutate(issue = case_when(code_check == "TRUE" & prius_name_check == "TRUE" ~ "ok",
                           code_check == "TRUE" & prius_name_check == "FALSE" ~ "name_issue",   #name_issue: in prius list Species is not equal to gbif scientific name
                           code_check == "FALSE" & prius_name_check == "TRUE" ~ "code_issue",   #code_issue: gbif codes differ between prius and union list
                           code_check == "FALSE" & prius_name_check == "FALSE" ~ "code_name_issue",   #code_name_issue: combination of two previous issues
                           is.na(code_check) & prius_name_check == "FALSE" ~ "name_issue")) %>%
  arrange(issue)
```

```{r me}
# Single out batches of issues
explore <- list_check %>% 
  select(Species,
         prius_scientificName,  union_scientificName, 
         prius_gbif_code, union_gbif_code,
         prius_status_check, union_status_check,
         issue) %>% 
  filter(issue == "name_issue")
View(explore)

rm(explore)
```

Initiating RadIUS based on the lessons learnt.

```{r}
radius <- prius %>%
  transmute(Groep,
            Species,
            GBIF_scientific_name,
            GBIF_code = if_else(GBIF_code == 3034613, 7978544, # Hydrocotyle
                                GBIF_code),
            Soort,
            Lijst)
```

As of 2025-08-07 : ook de andere issues nog eens goed checken.
Species = Naam_cf_verordening - kolom gewoon zo hernoemen?