```{r prep packages}
library(sf)
library(tidyverse)
library(vecsets)
library(colorRamps) # for continuous-scale plots (Matlab-like)

```

```{r prep memory}
rm(list = ls())
rm(list = setdiff(ls(), c("read_species_list", "ias.cube.raw")))
```

```{r source helpers}
source("./prius2/scripts/helper_functions.R")
```

```{r prep spatial}
# flanders (clean + grid)

map.flan.clean <- st_read("./prius/data/spatial/flanders_wgs84.geojson")
map.prov.clean <- st_read("./prius2/spatial/provincies2025.geojson") # !! updated
map.flan.grid <- st_read("./prius/data/spatial/grids/vla_1km.geojson")

# protected areas (clean + grid)
map.VRL.clean <- st_read("./prius2/spatial/vogelrichtlijn2025.geojson") # !! updated
map.HRL.clean <- st_read("./prius2/spatial/habitatrichtlijn2025.geojson") # !! updated

```

```{r get species list}
spp.list.raw <- read_species_list(file_path = "./prius2/temp/",
                                  prius_version = 2)
```

NOTE: the species cube is (too) heavy to push to the git. Therefore, at first use, it is downloaded to the local disk from Drive.

```{r update cube locally}
# fistools::download_gdrive_if_missing(gfileID = "154S4vww2G1f0vH7_jt1cZJXeUEmXlZaL",
#                                      destfile = "prius2/input/prius2_species_cube.csv",
#                                      update_always = TRUE,
#                                      email = Sys.getenv("email"))
```

```{r get cube data}
ias.cube.raw <- read_csv(file = "./prius2/input/prius2_species_cube.csv") %>% 
  as_tibble()

gc()
```

```{r exclude marine species}
# according to PrIUS philosophy, purely marine/brackish species are excluded (irrelevant)

irr <- read.csv("./prius2/temp/criteria_environment_v26-02-06.csv") %>% 
  mutate(irr = ifelse(zoetwater == 0 & terrestrisch == 0, 1, 0)) %>% 
  filter(irr == 1) %>% 
  print()

spp.list <- spp.list.raw %>% 
  filter(!GBIF_code %in% irr$GBIF_code)

ias.cube <- ias.cube.raw %>% 
  filter(!specieskey %in% irr$GBIF_code)
```

```{r set time frame}
time.frame.start <- 2015 # !! first year
time.frame.stop  <- 2025 # !! last year (included)
```

```{r wrangle ias.cube Flanders}
ias.cube <- ias.cube %>%
  filter(year >= time.frame.start &
           year <= time.frame.stop) %>%
  filter(specieskey %in% spp.list$GBIF_code) %>% 
  filter(eeacellcode %in% map.flan.grid$CELLCODE)

ias.cube.summ <- ias.cube %>%
  group_by(specieskey) %>% 
  summarise(unique_cells = n_distinct(eeacellcode))
```

# ABSENT / SPORADIC / PRESENT (Flanders)

```{r}
spp.list.metrics <- spp.list %>%
  select(soort, GBIF_canonicalName, GBIF_code) %>%
  left_join(ias.cube.summ, by = c("GBIF_code" = "specieskey"))
```

```{r absent}
# "Soorten die in geen enkel kalenderjaar zijn vastgesteld, beschouwen we als afwezig."
spp.list.metrics <- spp.list.metrics %>% 
  mutate(aanwezigheid = ifelse(is.na(unique_cells), "afwezig", NA))
```

```{r present}
# "Soorten die in elk van de vijf meest recente kalenderjaren zijn vastgesteld -> aanwezig"

present.time <- c(seq(time.frame.stop-4, time.frame.stop))

present.temp.L <- ias.cube %>% 
  group_by(species, specieskey, year) %>% 
  summarize(unique_cells = n_distinct(eeacellcode))
  # filter(year %in% present.time)

present.temp.W <- present.temp.L %>% 
  pivot_wider(values_from = unique_cells,
              names_from = year)

# order column names
ix  <- 3:ncol(present.temp.W)
present.temp.W  <- present.temp.W[ , c(1:2,
  ix[order(names(present.temp.W)[ix])]
  )]

present.temp.W <- present.temp.W %>% 
  mutate(aanwezigheid = if_else(if_all(as.character(c(
    time.frame.stop - 4, # prius-2 : year 2021
    time.frame.stop - 3,           # year 2022
    time.frame.stop - 2,           # year 2023
    time.frame.stop - 1,           # year 2024
    time.frame.stop)),             # year 2025
    ~ !is.na(.x)), "aanwezig", NA))

spp.list.metrics <- spp.list.metrics %>%
  left_join(present.temp.W[, c("specieskey", "aanwezigheid")],
            by = c("GBIF_code" = "specieskey"))
            
spp.list.metrics <- spp.list.metrics %>%
  mutate(aanwezigheid = coalesce(aanwezigheid.x, aanwezigheid.y)) %>%
  select(-aanwezigheid.x,
         -aanwezigheid.y)
```

```{r sporadic}
# "De overige soorten beschouwen we als sporadisch."

spp.list.metrics$aanwezigheid[is.na(spp.list.metrics$aanwezigheid)] = "sporadisch"
```

```{r write}
# tussentijds wegschrijven
take.out <- spp.list.raw[c("soort", "GBIF_canonicalName", "reg")] %>%
  left_join(spp.list.metrics[c("GBIF_canonicalName", "aanwezigheid")],
            by = "GBIF_canonicalName")
write.csv(take.out,
          paste0("./prius2/temp/criteria_aanwezigheid_", Sys.Date(), ".csv"),
          row.names = F)
```

# ABSENT / SPORADIC / PRESENT (neighbouring regions)

The same analysis is performed for the regions neighbouring Flanders (WA, FR, NL...). As to retain the focus in this script on Flanders, this neighbour-region-analysis is developed in a different script.

# NATURA 2000

```{r Calulate_N2000_occupancy_metrics}

#Load Flemish shapefiles with EEA grid cells for Natura 2000, Birds's directive, and Habitat's directive areas
eea_cells<-st_read(file.path("prius2", "spatial","utm1_bel_regions.geojson"))

map.N2K.grid<-eea_cells%>%
  filter(isFlandersN2000)

map.VRL.grid<-eea_cells%>%
  filter(isFlandersSPA)

map.HRL.grid<-eea_cells%>%
  filter(isFlandersSAC)


#Define the number of cells in Natura 2000, Birds's directive, and Habitat's directive areas
N_Natura2000 <- as.numeric(n_distinct(map.N2K.grid$CELLCODE))
N_VLR <- as.numeric(n_distinct(map.VRL.grid$CELLCODE))
N_HRL <- as.numeric(n_distinct(map.HRL.grid$CELLCODE))


#Prepare the species cube for calculations
species.cube <- ias.cube.raw %>% 
  dplyr::filter(!specieskey %in% irr$GBIF_code)%>%
  dplyr::filter(year >= 2015 &
           year <= 2025) %>%
  dplyr::filter(specieskey %in% spp.list$GBIF_code) %>% 
  dplyr::filter(eeacellcode %in% c(map.flan.grid$CELLCODE, map.N2K.grid$CELLCODE)) 


#Calculate a number of occupancy metrics for each species
df<- species.cube %>%
  left_join(spp.list %>%
              select(GBIF_code, soort), by = c("specieskey" = "GBIF_code")) %>%
  arrange(soort) %>%
  group_by(soort) %>%
  summarise(
    N_occ = as.numeric(n_distinct(eeacellcode)),
    
    N_occ_in_N2000 = as.numeric(n_distinct(eeacellcode[eeacellcode %in% map.N2K.grid$CELLCODE])),
    N_occ_in_VRL   = as.numeric(n_distinct(eeacellcode[eeacellcode %in% map.VRL.grid$CELLCODE])),
    N_occ_in_HRL   = as.numeric(n_distinct(eeacellcode[eeacellcode %in% map.HRL.grid$CELLCODE])),
    
    .groups = "drop"
  ) %>%
  mutate(
    share_in_N2000 = N_occ_in_N2000 / N_occ,
    share_in_HRL   = N_occ_in_HRL   / N_occ,
    share_in_VRL   = N_occ_in_VRL   / N_occ,
    
    `%_in_N2000` = round(100 * share_in_N2000, 1),
    `%_in_HRL`   = round(100 * share_in_HRL, 1),
    `%_in_VRL`   = round(100 * share_in_VRL, 1),
    
    share_of_N2000 = N_occ_in_N2000 / N_Natura2000,
    share_of_HRL   = N_occ_in_HRL   / N_HRL,
    share_of_VRL   = N_occ_in_VRL   / N_VLR,
    
    `‰_of_N2000` = round(1000 * share_of_N2000, 1),
    `‰_of_HRL`   = round(1000 * share_of_HRL, 1),
    `‰_of_VRL`   = round(1000 * share_of_VRL, 1)
  ) %>%
  rename(species = soort)%>%
  select(-c(N_occ_in_VRL, N_occ_in_HRL))%>%
  as.data.frame()
```

```{r}
write_csv(df, "./prius2/output/occupancy_Natura2000/occupancy_Natura2000.csv")
```

```{r}
# The only purpose of this chunck is to prepare the data for easy copy-paste into the PrIUS report
helper <- df %>% 
  left_join(spp.list, by = c("species" = "soort")) %>% 
  select(species, Species = species_wettelijke_naam,
         `%_in_N2000`, `%_in_HRL`, `%_in_VRL`,
         `‰_of_N2000`, `‰_of_HRL`, `‰_of_VRL`) %>% 
  arrange(Species)

for (soortje in unique(helper$Species)){
  
  A <- subset(helper, Species == soortje)
  B <- matrix(A[3:8], nrow = 3, ncol = 2, byrow = FALSE)

  # C <- as.data.frame(B)
  # D <- format(C, decimal.mark = ",", nsmall = 1)
  # lay-out goed, maar print helaas niet in de Console

  print(soortje)
  print(B, row.names = FALSE, right = FALSE)

  }
# Take into account the adhoc species (see elsewhere), viz. Aedes albopictus
```

# DISTRIBUTION MAPS

```{r maps}
# note : colour gradient (matlab-like) over years, as requested by ANB
# note : to save space, legend not included in graphs, but shown in separate fig.

spp.to.map <- subset(spp.list.metrics, aanwezigheid != "afwezig")
my.cols <- matlab.like(n_distinct(seq(time.frame.start, time.frame.stop)))

for (GBIF_canonicalName in spp.to.map$GBIF_canonicalName){
  
  print(GBIF_canonicalName)
  GBIF_code <- spp.to.map[spp.to.map$GBIF_canonicalName == GBIF_canonicalName, "GBIF_code"]
  print(GBIF_code)
  
  temp.ias.cube <- ias.cube %>%
    filter(specieskey %in% GBIF_code) %>%
    group_by(eeacellcode, specieskey) %>%
    summarise(year = as.integer(max(year, na.rm = TRUE)))
  
  temp.ias.cube.spec <- merge(map.flan.grid,
                              temp.ias.cube,
                              by.x = 'CELLCODE',
                              by.y = 'eeacellcode',
                              all.y = TRUE)
  
  temp_plot <- ggplot() +
    geom_sf(data = map.flan.clean, fill= "#f0f0f0", size = 0.2) +
    geom_sf(data = map.VRL.clean, colour= "#a4e98f", fill = "#a4e98f") + 
    geom_sf(data = map.HRL.clean, colour="#a4e98f", fill = "#a4e98f") +
    geom_sf(data = map.prov.clean, fill= NA, size = 0.2) +
    geom_sf(data = temp.ias.cube.spec,  aes(colour = year, fill = year),
            stroke = 0) + # only show filling, not lines
    scale_colour_gradientn(colors = my.cols,
                           limits = c(time.frame.start, time.frame.stop)) +
    scale_fill_gradientn(colors = my.cols,
                         limits = c(time.frame.start, time.frame.stop)) +
    theme_void() +
    theme(
      legend.position = "none",
      legend.title = element_blank()
      ) +
    coord_sf()
  
  ggsave(temp_plot, file=paste0(
    "./prius2/output/distribution_maps/", GBIF_canonicalName,".png"),
    width = 15.0, height = 5.1, # important for fluent incorporation in report
    units = "cm", dpi = 200)
}

rm(spp.to.map, my.cols)
```

```{r legend}
# one-off figure to extract the legend
# based on the last species of the above loop 

ggplot() +
  geom_sf(data = map.flan.clean, fill= "#f0f0f0", size = 0.2) +
  geom_sf(data = temp.ias.cube.spec,  aes(colour = year, fill = year),
          stroke = 0) +
  scale_colour_gradientn(colors = my.cols,
                         limits = c(time.frame.start, time.frame.stop),
                         breaks = seq(time.frame.start, time.frame.stop, by = 2)) +
  scale_fill_gradientn(colors = my.cols,
                       limits = c(time.frame.start, time.frame.stop),
                       breaks = seq(time.frame.start, time.frame.stop, by = 2)) +
  theme_void() +
  theme(
    legend.position = "bottom",
    legend.key.width = unit(2, "cm"),
    legend.title = element_blank()
    ) +
    coord_sf()

ggsave(file=paste0("./prius2/output/distribution_maps/",
                   "_legende.png"),
       width = 15.0, height = 5.1,
       units = "cm", dpi = 200)
```