```{r prep packages}
library(sf)
library(tidyverse)
library(vecsets)
library(colorRamps) # for continuous-scale plots (Matlab-like)

source("./prius2/scripts/helper_functions.R")
```

```{r prep memory}
rm(list = ls())
rm(list = setdiff(ls(), c("read_species_list", "ias.cube.raw")))
```

```{r prep spatial}
# flanders (clean + grid)

map.flan.clean <- st_read("./prius/data/spatial/flanders_wgs84.geojson")
map.prov.clean <- st_read("./prius2/spatial/provincies2025.geojson") # !! updated
map.flan.grid <- st_read("./prius/data/spatial/grids/vla_1km.geojson")

# protected areas (clean + grid)

map.VRL.clean <- st_read("./prius/data/spatial/ps_vglrl.geojson") # VRL = vogelrichtlijn
map.HRL.clean <- st_read("./prius/data/spatial/ps_hbtrl_deel.geojson") # HRL = habitatrichtlijn

map.N2K.raw <- read_tsv(
  "https://raw.githubusercontent.com/trias-project/indicators/master/data/interim/intersect_EEA_ref_grid_protected_areas.tsv",
  na = "")

map.N2K.grid <- map.N2K.raw %>% filter(natura2000 == TRUE)
map.SPA.grid <- map.N2K.raw %>% filter(spa == TRUE) # = VRL? hernoemen?
map.HRL.grid <- map.N2K.raw %>% filter(habitat == TRUE)

rm(map.N2K.raw)
```

# Species list

```{r get species list}
spp.list.raw <- read_species_list(file_path = "./prius2/temp/",
                                  prius_version = 2)
```

NOTE: the species cube is (too) heavy to push to the git. This will give an error when running the script. I needed to download it locally (making connection with zenodo within the script was too slow). So maybe best to download the most recent version of the species cube and put it in the folder data/input before running it.

```{r update cube locally}
fistools::download_gdrive_if_missing(gfileID = "154S4vww2G1f0vH7_jt1cZJXeUEmXlZaL",
                                     destfile = "prius2/input/prius2_species_cube.csv",
                                     update_always = TRUE,
                                     email = Sys.getenv("email"))
```

```{r get cube data}
ias.cube.raw <- read_csv(file = "./prius2/input/prius2_species_cube.csv") %>% 
  as_tibble()

gc()
```

```{r exclude marine species}
# according to PrIUS philosophy, purely marine/brackish species are excluded (irrelevant)

irr <- read.csv("./prius2/temp/criteria_environment_v26-01-22.csv") %>% 
  mutate(irr = ifelse(zoetwater == 0 & terrestrisch == 0, 1, 0)) %>% 
  filter(irr == 1) %>% 
  print()

spp.list <- spp.list.raw %>% 
  filter(!GBIF_code %in% irr$GBIF_code)

ias.cube <- ias.cube.raw %>% 
  filter(!specieskey %in% irr$GBIF_code)
```

```{r set time frame}
time.frame.start <- 2015 # !! first year
time.frame.stop  <- 2025 # !! last year (included)
```

```{r wrangle ias.cube}
ias.cube <- ias.cube %>%
  filter(year >= time.frame.start &
           year <= time.frame.stop) %>%
  filter(specieskey %in% spp.list$GBIF_code) %>% 
  filter(eeacellcode %in% map.flan.grid$CELLCODE)

ias.cube.summ <- ias.cube %>%
  group_by(specieskey) %>% 
  summarise(n_tot = sum(occurrences, na.rm = TRUE), 
            unique_cells = n_distinct(eeacellcode))
```

# ABSENT / SPORADIC / PRESENT

```{r}
spp.list.metrics <- spp.list %>%
  select(soort, GBIF_code) %>%
  left_join(ias.cube.summ, by = c("GBIF_code" = "specieskey"))

```

```{r absent}
spp.list.metrics <- spp.list.metrics %>% 
  mutate(aanwezigheid = ifelse(is.na(unique_cells), "afwezig", NA))
```

sporadic vs. present

```{r needed?}
cells_per_species <- ias.cube.summ %>%
  left_join(spp.list, by = c("specieskey" = "GBIF_code")) %>%
  select(specieskey, unique_cells, soort)

not_in_cube <- spp.list %>%
  left_join(ias.cube.summ, by = c("GBIF_code" = "specieskey")) %>%
  filter(is.na(n_tot))

write_csv(not_in_cube, "./prius2/output/taxonkeys_not_in_cube.csv")

in_cube <- spp.list %>%
  left_join(ias.cube.summ, by = c("GBIF_code" = "specieskey")) %>%
  filter(!is.na(n_tot))

not_in_cube$soort %in% in_cube$soort

List_species_present <- unique(in_cube$soort)

write_csv2(cells_per_species, "./prius2/output/cells_per_species.csv")
write(List_species_present, "./prius2/output/species_present.txt")
```

```{r}
keep_all <- data.frame(year = seq.int(from = 2015,
                                      to = 2025,
                                      by = 1)) ## Keep all species

keep_all$year <- as.factor(keep_all$year) ## For compatibility with below

for (species_name in unique(in_cube$soort)){
  print(species_name)
  GBIF_codes<- in_cube[which(in_cube$soort==species_name),]$GBIF_code
  print(GBIF_codes)
  IAS_Cube_temp <- ias.cube %>%
    filter(specieskey %in% GBIF_codes)
  print(IAS_Cube_temp)
  IAS_Cube_1km_spec <-IAS_Cube_temp %>% 
    left_join(map.flan.grid, by = c("eeacellcode" = "CELLCODE")) %>% 
    as.data.frame() %>%
    select(-geometry) %>%
    filter(occurrences > 0)
  IAS_Cube_sporadic <- IAS_Cube_1km_spec %>% 
  group_by(year)%>%
  summarize(n=n())
  
  IAS_Cube_sporadic$year <- as.factor(IAS_Cube_sporadic$year)
  
  data <- data.frame(year = seq.int(from = 2015,
                                      to = 2025,
                                      by = 1))
  
  data$year <- as.factor(data$year)
  
  x <- merge(data, IAS_Cube_sporadic)
  x <- replace(x, is.na(x), 0)
                  
  temp_plot<- ggplot(data=x, aes(x=year, y=n)) +
  geom_bar(stat="identity")+
  labs(y = expression ("Occupancy in"~km^2))+
  scale_x_discrete(drop = FALSE)
  
  # ggsave(temp_plot, file=paste0("./data/output/absent-sporadic-present/", species_name,".png"), dpi=200)
  
  keep_all <- keep_all %>% 
    left_join(x[,1:2], by = "year") %>% 
    rename(!!species_name:=n)  ## Geef kolom naam van soort (m.o.: gebruik !! := want variabele)
    
}
```

```{r summary across species}
showme <- t(keep_all)             ## Transpose
colnames(showme) <- showme[1,]    ## First row to column names...
showme <- showme[-1,]             ## ... and delete first row
showme <- as.data.frame(showme)

# Determine whether the species is absent, sporadic or present
# Absent: 0 occupied 1km2 cells in all years
# Present: more than 1km2 cells occupied 5 last years (2020-2025)
# Sporadic: everything else
showme <- showme %>%
  rownames_to_column(var = "species") %>%
  mutate(across(-species, as.numeric)) %>% 
  rowwise() %>%
  mutate(status = case_when(
    sum(c_across(`2015`:`2025`)) == 0 ~ "absent",
    sum(c_across(`2020`:`2025`) > 0) >= 5 ~ "present",
    TRUE ~ "sporadic"
  ))
  

write_csv(showme, "./prius2/output/absent-sporadic-present/absent-sporadic-present.csv")
```

# NATURA 2000

```{r}
df <- data.frame()

for (species_name in unique(in_cube$soort)){
  
  print(species_name)
  GBIF_codes <- in_cube[which(in_cube$soort==species_name),]$GBIF_code
  print(GBIF_codes)
  IAS_Cube_temp <- ias.cube %>%
    filter(specieskey %in% GBIF_codes)

  N_occ_spec <- as.numeric(n_distinct(IAS_Cube_temp$eeacellcode)) # nodig, want zelfde cel voor verschillende jaren
  N_Natura2000 <- as.numeric(n_distinct(map.N2K.grid$CELLCODE))
  N_spa <- as.numeric(n_distinct(map.SPA.grid$CELLCODE))
  N_habitat <- as.numeric(n_distinct(map.HRL.grid$CELLCODE))
  
  N_occ_spec_in_Natura2000 <- as.numeric(n_distinct(
    IAS_Cube_temp[which(IAS_Cube_temp$eeacellcode %in% map.N2K.grid$CELLCODE),]$eeacellcode))
  N_occ_spec_in_spa <- as.numeric(n_distinct(
    IAS_Cube_temp[which(IAS_Cube_temp$eeacellcode %in% map.SPA.grid$CELLCODE),]$eeacellcode))
  N_occ_spec_in_habitat <- as.numeric(n_distinct(
    IAS_Cube_temp[which(IAS_Cube_temp$eeacellcode %in% map.HRL.grid$CELLCODE),]$eeacellcode))
                  
df <- rbind(df, data.frame(species_name,
                           N_occ_spec,
                           N_occ_spec_in_Natura2000,
                           N_occ_spec_in_Natura2000/N_occ_spec,
                           N_occ_spec_in_habitat/N_occ_spec,
                           N_occ_spec_in_spa/N_occ_spec,
                           round(100*(N_occ_spec_in_Natura2000/N_occ_spec), digits = 1),  # Idem, as percent (%), for report
                           round(100*(N_occ_spec_in_habitat/N_occ_spec), digits = 1),     # Idem, as percent (%), for report
                           round(100*(N_occ_spec_in_spa/N_occ_spec), digits = 1),         # Idem, as percent (%), for report
                           N_occ_spec_in_Natura2000/N_Natura2000,
                           N_occ_spec_in_habitat/N_habitat,
                           N_occ_spec_in_spa/N_spa,
                           round(1000*(N_occ_spec_in_Natura2000/N_Natura2000), digits = 1), # Idem, as promille (‰), for report
                           round(1000*(N_occ_spec_in_habitat/N_habitat), digits = 1),       # Idem, as promille (‰), for report
                           round(1000*(N_occ_spec_in_spa/N_spa), digits = 1)))              # Idem, as promille (‰), for report
}

colnames(df) <- c("species",
                  
                  "N_occ",
                  "N_occ_in_N2000",
                  
                  "share_in_N2000",
                  "share_in_HRL",
                  "share_in_VRL",
                  
                  "%_in_N2000",
                  "%_in_HRL",
                  "%_in_VRL",
                  
                  "share_of_N2000",
                  "share_of_HRL",
                  "share_of_VRL",
                  
                  "‰_of_N2000",
                  "‰_of_HRL",
                  "‰_of_VRL")
```

```{r}
write_csv(df, "./prius2/output/occupancy_Natura2000/occupancy_Natura2000.csv")
```

```{r}
# The only purpose of this chunck is to prepare the data for easy copy-paste into the PrIUS report
helper <- df %>% 
  left_join(spp.list, by = c("species" = "soort")) %>% 
  select(species, Species = species_wettelijke_naam,
         `%_in_N2000`, `%_in_HRL`, `%_in_VRL`,
         `‰_of_N2000`, `‰_of_HRL`, `‰_of_VRL`) %>% 
  arrange(Species)

for (soortje in unique(helper$Species)){
  
  A <- subset(helper, Species == soortje)
  B <- matrix(A[3:8], nrow = 3, ncol = 2, byrow = FALSE)

  # C <- as.data.frame(B)
  # D <- format(C, decimal.mark = ",", nsmall = 1)
  # lay-out goed, maar print helaas niet in de Console

  print(soortje)
  print(B, row.names = FALSE, right = FALSE)

  }
# Take into account the adhoc species (see elsewhere), viz. Aedes albopictus
```

# DISTRIBUTION MAPS

```{r maps}
# note : colour gradient (matlab-like) over years, as requested by ANB
# note : to save space, legend not included in graphs, but shown in separate fig.

my.cols <- matlab.like(n_distinct(seq(time.frame.start, time.frame.stop)))

for (species_name in unique(in_cube$soort)){
  print(species_name)
  GBIF_codes <- in_cube[which(in_cube$soort==species_name),]$GBIF_code
  print(GBIF_codes)
  IAS_Cube_temp <- ias.cube %>%
    filter(specieskey %in% GBIF_codes) %>% 
    group_by(eeacellcode, specieskey) %>%
    summarise(year = as.integer(max(year, na.rm = TRUE)))
  print(IAS_Cube_temp)
  IAS_Cube_1km_spec <- merge(map.flan.grid,
                             IAS_Cube_temp,
                             by.x = 'CELLCODE',
                             by.y = 'eeacellcode',
                             all.y = TRUE)
  
  temp_plot <- ggplot() +
    geom_sf(data = map.flan.clean, fill= "#f0f0f0", size = 0.2) +
    geom_sf(data= map.VRL.clean, colour= "#a4e98f", fill="#a4e98f") + 
    geom_sf(data= map.HRL.clean, colour="#a4e98f", fill="#a4e98f") +
    geom_sf(data = map.prov.clean, fill= NA, size=0.2) +
    geom_sf(data = IAS_Cube_1km_spec,  aes(colour = year, fill = year),
            stroke = 0) + # only show filling, not lines
    scale_colour_gradientn(colors = my.cols,
                         limits = c(time.frame.start, time.frame.stop)) +
    scale_fill_gradientn(colors = my.cols,
                         limits = c(time.frame.start, time.frame.stop)) +
    theme_void() +
    theme(
      legend.position = "none",
      legend.title = element_blank()
      ) +
    coord_sf()
  
  ggsave(temp_plot, file=paste0(
    "./prius2/output/distribution_maps/", species_name,".png"),
    width = 15.0, height = 5.1, # important for fluent incorporation in report
    units = "cm", dpi = 200)
}
```

```{r legend}
# one-off figure to extract the legend
# based on the last species of the above loop 

ggplot() +
  geom_sf(data = map.flan.clean, fill= "#f0f0f0", size = 0.2) +
  geom_sf(data = IAS_Cube_1km_spec,  aes(colour = year, fill = year),
          stroke = 0) +
  scale_colour_gradientn(colors = my.cols,
                         limits = c(time.frame.start, time.frame.stop),
                         breaks = seq(time.frame.start, time.frame.stop, by = 2)) +
  scale_fill_gradientn(colors = my.cols,
                       limits = c(time.frame.start, time.frame.stop),
                       breaks = seq(time.frame.start, time.frame.stop, by = 2)) +
  theme_void() +
  theme(
    legend.position = "bottom",
    legend.key.width = unit(2, "cm"),
    legend.title = element_blank()
    ) +
    coord_sf()

ggsave(file=paste0("./prius2/output/distribution_maps/",
                   "_legende.png"),
       width = 15.0, height = 5.1,
       units = "cm", dpi = 200)
```