---
title: "Emerging status: pre-processing"
author:
- Damiano Oldoni
- Toon Van Daele
- Tim Adriaens
- Soria Delva
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_depth: 3
    toc_float: true
    number_sections: true
editor_options: 
  chunk_output_type: console
---

This document describes the pre-processing of occurrence data of alien species. The outputs of this pipeline will be used as input for GAM modelling.

# Setup

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, tidy='styler')
```

Load libraries:

```{r load_libraries, message = FALSE}
library(tidyverse) # To do data science
library(tidylog) # To provide feedback on dplyr functions
library(here) # To find files
library(lubridate) # To work with dates
library(rgbif) # To get taxa from publisehd unified checklist
```

# Get data

## Occurrence data

We read the occurrence cube of the PrIUS II taxa, be_alientaxa_cube, generated by Sander Devisscher.
This was generated on 04/02/2026.


```{r read_data_cube, message = FALSE, warning = FALSE}
df <- read_csv(
  file =here::here("prj_plantes_exotiques", "data", "input", "be_alientaxa_cube.csv"),
  col_types = cols(
    year = col_double(),
    eea_cell_code = col_character(),
    taxonKey = col_double(),
    obs = col_double(),
    min_coord_uncertainty = col_double()
  ),
  na = ""
)%>%
 tidylog::select(-min_coord_uncertainty)%>% #Remove columns we don't need
 tidylog::filter(!is.na(eea_cell_code)) #Remove data not assigned to any cell code
```


## Baseline (class) data

We read the occurrence cube of the respective classes of the PrIUS II taxa, be_classes_cube, generated by Sander Devisscher.
This was generated on 04/02/2026.

```{r get_data_baseline}
df_bl <- read_csv(
  file = here::here("prj_plantes_exotiques", "data", "input", "be_classes_cube.csv"),
  col_types = cols(
    year = col_double(),
    eea_cell_code = col_character(),
    classKey = col_double(),
    cobs = col_double(),
    min_coord_uncertainty = col_double()
  ),
  na = ""
)%>%
 tidylog::select(-min_coord_uncertainty)%>% #Remove columns we don't need
 tidylog::filter(!is.na(eea_cell_code))%>% #Remove data not assigned to any cell code
 tidylog::filter(!is.na(classKey)) #Remove data with classKey NA are these are related to IAS species not linked to any class
```


## Checklist data

We read input data based on [Global Register of Introduced and Invasive Species - Belgium](https://www.gbif.org/dataset/6d9e952f-948c-4483-9807-575348147c7e), typically referred to as the _unified checklist_. These data contain taxonomic information as well as distribution, description and species profiles. We limit to plant taxa introduced in Belgium (`locationId` : `ISO_3166:BE`). Here we use the GRIIS checklist generated using the updated version of the TrIAS indicators script 01_get_data_input_checklist_indicators.Rmd.The script was updated so that synonyms were converted to their accepted taxa and species Sedum pallidum M. Bieb was added manually (for now).

```{r unified_checklist_tidy_df}
taxa_df <-
  read_tsv(here::here("prj_plantes_exotiques", "data", "input", "data_input_checklist_indicators.tsv"),
    na = "",
    guess_max = 5000
  ) %>%
  tidylog::filter(locationId == "ISO_3166:BE", #Only keep species in Belgium
                  kingdom=="Plantae") #Only keep plant species
```

## Grid cells in Belgium

```{r read_regions_info}
gpkg_path <- here::here("prj_plantes_exotiques","data", "input", "utm1_bel_with_regions.gpkg")
eea_grid_regions <- sf::st_read(gpkg_path)
```

## Grid cells intersecting protected areas

We get information about the intersection of grid cells with Belgian protected areas by reading file `intersect_EEA_ref_grid_protected_areas.tsv` in `data/output`:

```{r read_protected_areas}
df_prot_areas <- readr::read_tsv(
  here::here(
    "prj_plantes_exotiques",
    "data",
    "input",
    "intersect_EEA_ref_grid_protected_areas.tsv"
  ),
  na = ""
)%>%
  tidylog::select( #Only keep columns we are interested in
    CELLCODE,
    natura2000
  )
```


#  Pre-processing

## Add taxonomy 

For better interpretation of the data, we retrieve scientific names (without authorship) of the species whose data we are going to analyze. We also retrieve the class of each alien species to link occurrence data with the related baseline data. This step requires an internet connection and can take long time as we are querying the GBIF API:

```{r get_canonical_names}
#Get taxon keys from the cube
taxon_key <-
  df %>%
  tidylog::distinct(taxonKey) %>%
  pull()

#Extract taxonomic information of these respective taxonkeys
spec_names <- purrr::map_df(
  taxon_key,
  function(k) {
    name_usage(key = k)$data
  }
) %>%
  tidylog::select(
    taxonKey = key,
    canonicalName,
    scientificName,
    kingdomKey, 
    classKey,
    kingdom,
    class
  ) %>%
  tidylog::mutate(canonicalName = ifelse(
    is.na(canonicalName), scientificName, canonicalName
  ))
```

In the rare case multiple taxa share the same scientific name, authorship is added:
  
```{r add_authorship_if_needed}
spec_names <-
  spec_names %>%
  tidylog::group_by(canonicalName) %>%
  tidylog::add_tally() %>%
  tidylog::ungroup() %>%
  tidylog::mutate(canonicalName = if_else(n > 1,
                                          scientificName,
                                          canonicalName)) %>%
  tidylog::select(-c(n, scientificName))
```


## Apply temporal and spatial restraints

Select alien species on the GRIIS checklist that are introduced after 1950 (`year_of_introduction`: 1950) or whose date of introduction is not known and add their taxonomic info.

```{r select_aliens_after_year_of_introduction}
year_of_introduction <- 1950

recent_alien_species <-
  taxa_df %>%
  # Lower limit on date of first introduction if present
  tidylog::filter(first_observed >= year_of_introduction | is.na(first_observed)) %>%
  # Remove duplicates due to other columns as pathways
  tidylog::distinct(nubKey, first_observed, last_observed) %>%
  # Get classKey info and filter out taxa without data
  tidylog::inner_join(spec_names, by = c("nubKey" = "taxonKey")) %>%
  tidylog::distinct(
    taxonKey = nubKey,
    .data$canonicalName,
    .data$first_observed,
    .data$last_observed,
    .data$class,
    .data$kingdom,
    .data$classKey,
    .data$kingdomKey
  )
```

Number of taxa recently introduced or whose introduction year is not specified:

```{r n_recent_taxa}
nrow(recent_alien_species)
```

List of alien species excluded as first introduction date is earlier than 1950:

```{r recent_alien_species}
old_introductions_taxa <-
  taxa_df %>%
  tidylog::filter(first_observed < year_of_introduction) %>%
  tidylog::anti_join(recent_alien_species %>%
    tidylog::select(taxonKey),
  by = c("nubKey" = "taxonKey")
  ) %>%
  tidylog::distinct(
    nubKey,
    first_observed,
    last_observed
  ) %>%
  tidylog::inner_join(spec_names, by = c("nubKey" = "taxonKey")) %>%
  tidylog::select(
    taxonKey = nubKey,
    canonicalName,
    first_observed,
    last_observed,
    class,
    kingdom,
    everything()
  )

old_introductions_taxa
```

We save them in file `taxa_introduced_in_BE_before_1950.tsv` in  `data/output`:

```{r save_old_introduced_taxa_as_tsv_file}
write_tsv(old_introductions_taxa,
  here::here(
    "prj_plantes_exotiques",
    "data",
    "output",
    paste0(
      "taxa_introduced_in_BE_before_",
      year_of_introduction,
      ".tsv"
    )
  ),
  na = ""
)
```

We also limit the time series analysis on observations from 1950, start of state of invasion researches:

```{r define_year_range}
first_year <- 2000
```

Apply temporal restraint to both occurrence data:

```{r apply_restraints_time_space_occ_data}
df <-
  df %>%
  tidylog::filter(taxonKey %in% recent_alien_species$taxonKey) %>% # Only keep plant taxa first observed in 1950 or later
  tidylog::filter(year > first_year) #Only keep occurrence data from after 1950
```

and baseline data:

```{r apply_restraints_time_space_baseline_data}
df_bl <-
  df_bl %>%
  tidylog::filter(year > first_year)
```

## Add information about presence in protected areas

We add whether the grid cell intersects any of the Natura2000 Belgian protected areas.

```{r add_natura2000_col}
df <-
  df %>%
  tidylog::left_join(df_prot_areas,
    by = c("eea_cell_code" = "CELLCODE")
  )
```

We do the same for baseline data:

```{r add_natura2000_col_baseline}
df_bl <-
  df_bl %>%
  tidylog::left_join(df_prot_areas,
    by = c("eea_cell_code" = "CELLCODE")
  )

```


## Create time series

Here we create a dataset with, for each taxonKey, the cells in which this species has been observed as well as the first year the species has been observed anywhere in Belgium.

For each species, define cells with at least one observation:

```{r distinct_cells_taxa}
df_cc <- 
  df %>%
  tidylog::group_by(taxonKey) %>%
  tidylog::distinct(eea_cell_code) %>%
  tidylog::ungroup()
```

For each species, identify the first year with at least one observation:

```{r begin_year_per_species}
df_begin_year <- 
  df %>%
  tidylog::group_by(taxonKey) %>%
  tidylog::summarize(begin_year = min(year))
```

For each species, combine `begin_year` and unique `eea_cell_code` as found above: 

```{r combine_begin_year_cells_taxa}
df_cc <- 
  df_cc %>%
  tidylog::left_join(df_begin_year, by = "taxonKey") %>%
  tidylog::select(taxonKey, begin_year, eea_cell_code)
```


For each cell (`eea_cell_code`) and species (`taxonKey`) we can now create a time series:

```{r make_ts, cache=TRUE}
make_time_series <- function(eea_cell_code, taxonKey, begin_year, last_year ) {
  expand_grid(eea_cell_code = eea_cell_code,
              taxonKey = taxonKey,
              year = seq(from = begin_year, to = last_year))

}

# create timeseries slots
df_ts <- purrr::pmap_dfr(df_cc, 
                         .f = make_time_series, 
                         last_year = year(Sys.Date())
)

## Add data

# Add for each taxonkey in each year and in each cell the number of observations (obs)
df_ts <- 
  df_ts %>%
  tidylog::left_join(
    df %>% tidylog::select(taxonKey,
                           year,
                           eea_cell_code,
                           obs), 
                     by = c("taxonKey", "year", "eea_cell_code"))

# Add membership to protected areas
df_ts <- 
  df_ts %>%
  tidylog::left_join(df_prot_areas %>% 
                       tidylog::select(CELLCODE, natura2000),
                     by = c("eea_cell_code" = "CELLCODE"))

# Add regional information
df_ts <- 
  df_ts %>%
  tidylog::left_join(eea_grid_regions %>% 
                       as.data.frame() %>%
                       tidylog::select(CELLCODE, 
                                       isFlanders, 
                                       isWallonia, 
                                       isBrussels),
                     by = c("eea_cell_code" = "CELLCODE"))

# Add classKey
df_ts <- 
  df_ts %>%
  tidylog::left_join(spec_names %>% tidylog::select(taxonKey, classKey), 
                     by = "taxonKey")
```

## Research effort correction

To correct the effect of research effort of an alien speies, we calculate the number of observations at class level excluding the observations of the alien species itself:

```{r correct_cobs_research_effort_bias}
# Add baseline data (at class level) diminished by obs of specific alien taxon
df_ts <- 
  df_ts %>%
  tidylog::left_join(df_bl %>% 
                       tidylog::select(year, eea_cell_code, classKey, cobs),
                     by = c("year", "eea_cell_code", "classKey")) %>%
  tidylog::mutate(cobs = cobs - obs)

# replace NAs with 0
df_ts <-
  df_ts %>%
  tidylog::replace_na(list(cobs = 0, obs = 0))
```

Add column for presence (1) or absence (0)

```{r add_presence_absence}
df_ts <- 
  df_ts %>%
  tidylog::mutate(pa_cobs = if_else(cobs > 0, 1, 0),
                  pa_obs = if_else(obs > 0, 1, 0))
```

Arrange order columns:

```{r set_order_columns}
df_ts <-
  df_ts %>%
  tidylog::relocate(taxonKey, 
                    year, 
                    eea_cell_code, 
                    obs, 
                    pa_obs, 
                    cobs, 
                    pa_cobs,
                    classKey,
                    isFlanders,
                    isWallonia,
                    isBrussels,
                    natura2000)

#Only keep data for Natura2000 sites in Wallonia
df_ts<-df_ts%>%
  filter(isWallonia & natura2000)

```


Filter out data that do actually belong to Natura2000 sites that are not under Walloon jurisdiction. These are cells that overlap with both Flanders and Wallonia but that belong to Natura2000 sites of Flanders.

```{r remove_cells_not_in_walloonNatura2000_sites}
cells_in_walloon_protected_areas <- read_tsv(
  file = here::here("prj_plantes_exotiques","data", "interim", "cells_of_Walloon_protected_areas.tsv"),
  na = ""
)

df_ts<-df_ts%>%
  left_join(cells_in_walloon_protected_areas, by=c("eea_cell_code"="CELLCODE"))%>%
  filter(!is.na(SITECODE))

```

# Save data

## Save timeseries and taxonomic metadata

Save `df_ts` as it will be used as start point of modelling pipelines:

```{r save_data_as_tsv}
write_tsv(df_ts, 
          file = here::here("prj_plantes_exotiques","data", "interim", "df_timeseries2000.tsv"), 
          na = ""
)
```

We save also the taxonomic information:

```{r save_spec_names}
write_tsv(spec_names,
          file = here::here("prj_plantes_exotiques","data", "interim", "timeseries_taxonomic_info2000.tsv"), na = ""
)
```
