---
title: "Get cells of Belgian EEA reference grid containing protected areas"
author: 
  - Damiano Oldoni
  - Tim Adriaens
  - Soria Delva
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_depth: 3
    toc_float: true
    number_sections: true
editor_options: 
  chunk_output_type: console
---

# Goal

Given the EEA reference grid of Belgium at 1km resolution and the Flemish Natura2000 protected areas, this document assigns cells to their respective Flemish Natura2000 site, or to Flanders in general, if applicable.

# Setup 

```{r load_libraries, message=FALSE, warning=FALSE}
library(tidyverse) 
library(tidylog)
library(here) 
library(sf)
library(mapview)
```

# Read input data

## Import EEA reference grid

We import UTM grid data of Belgium at 1 by 1 km resolution. All grids have CRS (Coordinate Reference System) equal to _Belge Lambert 1972_:
  
```{r import_grids_of_BE_from_EEA}
be_grid <- st_read(here::here("prius2","spatial", "utm1_bel"))
```

## Read FLemish Natura2000 protected areas 

We read the shapefiles for the habitat directive areas and the bird directive areas

```{r protected_areas}
habitatrichtlijn <- st_read(file.path("prius2", "spatial", "habitatrichtlijn2025.geojson"))
vogelrichtlijn <- st_read(file.path("prius2", "spatial", "vogelrichtlijn2025.geojson"))

```

Convert to one big polygon and make sure crs matches that of be_grid

```{r crs_comparison}
habitatrichtlijn <- habitatrichtlijn %>%
  st_union() %>%
  st_transform(st_crs(be_grid)) %>%
  st_as_sf() %>%
  dplyr::mutate(name = "Habitatrichtlijn_Flanders")%>%
  dplyr::rename(geometry = x)

vogelrichtlijn <- vogelrichtlijn %>%
  st_union() %>%
  st_transform(st_crs(be_grid)) %>%
  st_as_sf() %>%
  dplyr::mutate(name = "vogelrichtlijn_Flanders")%>%
  dplyr::rename(geometry = x)
```

## Read shapefile of Flanders

```{r Flanders}
Flanders<- st_read("./prius/data/spatial/flanders_wgs84.geojson")%>%
   st_transform(st_crs(be_grid))
  
```

## Add regional information to UTM1 cells

In this section we add columns to know whether a cell belongs to Flanders or not and whether it belongs to a Flemish protected area 

```{r}
be_grid$isFlanders <-st_intersects(be_grid, Flanders, sparse = FALSE)[, 1]
be_grid$isFlandersSAC <- st_intersects(be_grid, habitatrichtlijn, sparse = FALSE)[, 1] #Habitatrichtlijngebieden
be_grid$isFlandersSPA <-st_intersects(be_grid, vogelrichtlijn, sparse = FALSE)[, 1] #Vogelrichtlijngebieden
be_grid <- be_grid%>%
  dplyr::mutate(isFlandersN2000 = isFlandersSAC | isFlandersSPA)
```


## Visualize to check

```{r Visualize Natura2000 cells}
n2000<-dplyr::filter(be_grid, isFlandersN2000)
ggplot()+
  geom_sf(data=Flanders, fill="grey")+
  geom_sf(data=n2000, fill="blue")+
  geom_sf(data=vogelrichtlijn, fill="green")+
  geom_sf(data=habitatrichtlijn, fill="orange")+
  theme_bw()


  mapview(Flanders, col.regions = "grey", alpha = 0.4, layer.name = "Flanders")+
  mapview(n2000, col.regions = "blue", alpha = 0.4, layer.name = "Natura2000")+
  mapview(vogelrichtlijn, col.regions="red", alpha = 0.4, layer.name = "SPA")+
  mapview(habitatrichtlijn,col.regions="orange", alpha = 0.4, layer.name = "SAC")

```

##Export be_grid with region info
```{r Store be_grid}
st_write(be_grid, file.path("prius2", "spatial", "utm1_bel_regions.geojson"), driver = "GeoJSON")

```

